Computing Entropies With Nested Sampling

### 2017-06-20 paper/ms.tex
\xref

\renewcommand{\d}{\boldsymbol{d}}
\newcommand{\todo}{\color{orange} \bf}
\newcommand{\query}{\color{green} \bf}
\newcommand{\x}{\boldsymbol{\theta}}
\newcommand{\bphi}{\boldsymbol{\phi}}
\newcommand{\boldeta}{\boldsymbol{\eta}}
\newcommand{\n}{\boldsymbol{\eta}}
\newcommand{\depth}{(\textnormal{depth})}

@@

The idea is to generate a reference particle $\x_{\rm ref}$
from the distribution $p(\x)$
whose entropy is required. Then, a Nested Sampling run evolves a set of
particles $\{\x\}_{i=1}^N$, initially representing $p(\x)$, is evolved
towards $\x_{\rm ref}$ in order to measure the log-probability near
$\x_{\rm ref}$.

@@

\While{more iterations desired}
    \State $k \leftarrow 0$
            \Comment{Initialise counter}
    \State Generate $\x_{\rm ref}$ from $p(\x)$
            \Comment{Generate a reference point}
\color{green}
    \State Generate $\left\{\x_i\right\}_{i=1}^N$ from $p(\x)$
            \Comment{Generate initial NS particles}
    \State Calculate $d_i \leftarrow d(\x_i; \x_{\rm ref})$ for all $i$
            \Comment{Calculate distance of each particle from the reference
                     point}
    \State $i^* \leftarrow \textnormal{argmin}(\lambda i \rightarrow d_i)$
            \Comment{Find the worst particle (greatest distance)}
    \State $d_{\rm max} \leftarrow d_{i^*}$
            \Comment{Find the greatest distance}
    \While{$d_{\rm max} > L$}
        \State Replace $\x_{i^*}$ with $\x_{\rm new}$ from
               $p\left(\x | (d(\x) < d_{\rm max})\right)$
            \Comment{Replace worst particle}
        \State Calculate $d_{i^*} \leftarrow d(\x_{\rm new}; \x_{\rm ref})$
            \Comment{Calculate distance of new particle from reference point}
        \State $i^* \leftarrow \textnormal{argmin}(\lambda i \rightarrow d_i)$
                \Comment{Find the worst particle (greatest distance)}
        \State $d_{\rm max} \leftarrow d_{i^*}$
                \Comment{Find the greatest distance}
        \State $k \leftarrow k+1$
            \Comment{Increment counter $k$}
    \EndWhile
\color{black}
    \State $\widehat{\mathbf{h}} \leftarrow \widehat{\mathbf{h}} + [k/N]$
            \Comment{Append latest estimate to results}
\EndWhile
\State $\widehat{h_{\rm final}} \leftarrow \frac{1}{\rm num\_iterations} \sum \widehat{\mathbf{h}}$
            \Comment{Average results}
\end{algorithmic}
\caption{Calculates $H(\x) = -\int p(\x) \log p(\x) \, d\x$
         without evaluating $\log p(\x)$. Technically, what it actually
         estimates is $\int f(\x) \int f(\x')
            \log \left[ P(d(\x', \x) < L | \x) \right]
                        \, d\x' \, d\x$, the expected value of the
        log-probability of a small region near $\x$.
        The part highlighted in green is standard Nested Sampling
        with quasi-prior $p(\x)$ and quasi-likelihood given by minus a
        distance function $d(\x', \x)$.
        \label{alg:algorithm}}
\end{algorithm}

@@

Continuous hypothesis spaces are commonly used in practice. Suppose
$\x$ has probability density $f(\x)$. The algorithm actually estimates
the `depth', or minus the expected value of the log-probability near $x$:
\begin{align}
h &= -\int f(\x) \int f(\x')
            \log \left[ P(d(\x', \x) < L | \x) \right]
                        \, d\x' \, d\x.
\end{align}
for some tolerance $L$.

If we actually want the differential
entropy
\begin{align}
H(\x) &= -\int f(x) \log f(x) \, dx
\end{align}
we can use the fact that density equals mass divided
by volume. Assume $L$ is small, so that
\begin{align}
P(d(\x', \x) < \epsilon | \x)
    &\approx
    f(\x) \int_{d(\x', \x) < L} \, d\x'\\
    &= f(\x) \, V
\end{align}
where $V$ is the volume of the region where $d(\x', \x) < L$.

If the distance function $d(\x', \x)$ is chosen to be Euclidean,
the constraint $d(\x', \x) < L$ corresponds to a ball in the space
of possible $\x'$ values.
The log-volume of a ball of radius $L$ in $n$ dimensions is
\begin{align}
\log V(\epsilon; n) &= \frac{n}{2}\log \pi
                        - \log \Gamma\left(\frac{n}{2} + 1\right)
                        + n \log L,
\end{align}
and provides the conversion factor between the differential entropy and
the entropy actually computed from the algorithm.

>>>>

\renewcommand{\d}{\boldsymbol{d}}
\newcommand{\todo}{\color{orange} \bf}
\newcommand{\query}{\color{green} \bf}
\newcommand{\x}{\boldsymbol{\theta}}
\newcommand{\bphi}{\boldsymbol{\phi}}
\newcommand{\boldeta}{\boldsymbol{\eta}}
\newcommand{\n}{\boldsymbol{\eta}}
\newcommand{\depth}{(\textnormal{depth})}
\newcommand{\xref}{\x_{\rm ref}}

@@

The idea is to generate a reference particle $\xref$
from the distribution $p(\x)$
whose entropy is required. Then, a Nested Sampling run evolves a set of
particles $\{\x\}_{i=1}^N$, initially representing $p(\x)$, is evolved
towards $\xref$ in order to measure the log-probability near
$\xref$. Nearness is defined using a distance function
$d(\x; \xref)$, and the number of NS iterations taken to reduce
the distance to below some threshold $L$ provides an unbiased estimate of
\begin{align}
   -\log \left[ P(d(\x; \xref) < L) \right].
\end{align}

@@

\While{more iterations desired}
    \State $k \leftarrow 0$
            \Comment{Initialise counter}
    \State Generate $\xref$ from $p(\x)$
            \Comment{Generate a reference point}
\color{green}
    \State Generate $\left\{\x_i\right\}_{i=1}^N$ from $p(\x)$
            \Comment{Generate initial NS particles}
    \State Calculate $d_i \leftarrow d(\x_i; \xref)$ for all $i$
            \Comment{Calculate distance of each particle from the reference
                     point}
    \State $i^* \leftarrow \textnormal{argmin}(\lambda i \rightarrow d_i)$
            \Comment{Find the worst particle (greatest distance)}
    \State $d_{\rm max} \leftarrow d_{i^*}$
            \Comment{Find the greatest distance}
    \While{$d_{\rm max} > L$}
        \State Replace $\x_{i^*}$ with $\x_{\rm new}$ from
               $p\left(\x | (d(\x) < d_{\rm max})\right)$
            \Comment{Replace worst particle}
        \State Calculate $d_{i^*} \leftarrow d(\x_{\rm new}; \xref)$
            \Comment{Calculate distance of new particle from reference point}
        \State $i^* \leftarrow \textnormal{argmin}(\lambda i \rightarrow d_i)$
                \Comment{Find the worst particle (greatest distance)}
        \State $d_{\rm max} \leftarrow d_{i^*}$
                \Comment{Find the greatest distance}
        \State $k \leftarrow k+1$
            \Comment{Increment counter $k$}
    \EndWhile
\color{black}
    \State $\widehat{\mathbf{h}} \leftarrow \widehat{\mathbf{h}} + [k/N]$
            \Comment{Append latest estimate to results}
\EndWhile
\State $\widehat{h_{\rm final}} \leftarrow \frac{1}{\rm num\_iterations} \sum \widehat{\mathbf{h}}$
            \Comment{Average results}
\end{algorithmic}
\caption{Calculates $H(\x) = -\int p(\x) \log p(\x) \, d\x$
         without evaluating $\log p(\x)$. Technically, what it actually
         estimates is $\int f(\x) \int f(\xref)
            \log \left[ P(d(\xref, \x) < L | \x) \right]
                        \, d\xref \, d\x$, the expected value of the
        log-probability of a small region near $\x$.
        The part highlighted in green is standard Nested Sampling
        with quasi-prior $p(\x)$ and quasi-likelihood given by minus a
        distance function $d(\xref, \x)$.
        \label{alg:algorithm}}
\end{algorithm}

@@

Continuous hypothesis spaces are commonly used in practice. Suppose
$\x$ has probability density $f(\x)$. The algorithm actually estimates
the `depth', or minus the expected value of the log-probability near $x$:
\begin{align}
h &= -\int f(\x) \int f(\xref)
            \log \left[ P(d(\xref, \x) < L | \x) \right]
                        \, d\xref \, d\x.
\end{align}
for some tolerance $L$.

If we actually want the differential
entropy
\begin{align}
H(\x) &= -\int f(x) \log f(x) \, dx
\end{align}
we can use the fact that density equals mass divided
by volume. Assume $L$ is small, so that
\begin{align}
P(d(\xref, \x) < \epsilon | \x)
    &\approx
    f(\x) \int_{d(\xref, \x) < L} \, d\xref\\
    &= f(\x) \, V
\end{align}
where $V$ is the volume of the region where $d(\xref, \x) < L$.

If the distance function $d(\xref, \x)$ is chosen to be Euclidean,
the constraint $d(\xref, \x) < L$ corresponds to a ball in the space
of possible $\xref$ values.
The log-volume of a ball of radius $L$ in $n$ dimensions is
\begin{align}
\log V(\epsilon; n) &= \frac{n}{2}\log \pi
                        - \log \Gamma\left(\frac{n}{2} + 1\right)
                        + n \log L,
\end{align}
and provides the conversion factor between the differential entropy and
the entropy actually computed from the algorithm.
